---
title: "Accelerate Your Workflow: Introduction to R"
author: "Thomas Sharpe, Senior Data Scientist - Idaho State University"
date: "14 June 2024"
output:
  html_document:
    toc: true
    toc_float: true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = TRUE)
```

# Challenge and Oppurtunity

Increasing demand for data coupled with flat or perhaps strained budgets

Chance to add a ton of value in the middle to long-term through acceleration of processes

R is a great option to achieve this!

# My Experience

* Consulting/Market Research (Two Years)
  + Junior Analyst Intern
  + Senior Project Manager
* College of Southern Idaho (~Seven Years)
  + Research Analyst
  + Senior Research Analyst
  + Director of Analytics and Institutional Research
* Idaho State University (Current)
  + Associate Director of Institutional Research
  + Senior Data Scientist (Current)
  
  
# Objective

Brief introduction to core R functionality such as:

* Base R
* Reading data
* Wrangling Data
* Summarizing Data
* Visualizing data
* Recap with an example, preparing to manually entering IPEDS Financial Aid. (If time)

# Base R

## Calculator

CTRL+ALT+I: R Code Chunk Shortcut

```{r}

#Comments
#CTRL+Enter to run an individual line of code
#CTRL+SHIFT+ENTER to run entire code chunk
2+2
2*100

```

## Variables

```{r}

#Both work
x <- 4
x = 5

x*3
#X*3 #R is case sensitive!


```

## Accessing your files

```{r}

#Your active working directory within your file system on your machine.
getwd()
#setwd() #if you would like to change it

dRetain <- read.csv('Retention.csv')
rm(dRetain)

#Default copy/paste: C:\Users\Thomas Sharpe\Desktop\AIHEC_IR24\IntroR
#\ is an escape character so you either need to switch it to / or \\ 
dRetain <- read.csv('C:\\Users\\Thomas Sharpe\\Desktop\\AIHEC_IR24\\IntroR\\Retention.csv')
dFAFSA <- read.csv('C:\\Users\\Thomas Sharpe\\Desktop\\AIHEC_IR24\\IntroR\\FAFSA.csv')

```

# Exploring Data

## Base R

```{r}

length(dRetain$STUDENT_ID)

length(unique(dRetain$FISCAL_YEAR))

dRetain$RETAIN_CODE <- ifelse(dRetain$RETAINED == 'Y',1,0)

sum(dRetain$RETAIN_CODE)

sum(dRetain$RETAIN_CODE) / length(dRetain$STUDENT_ID)

#dRetain

```

## Packages (The Good Stuff!)

How had female retention rate changes over-time?

```{r}

#install.packages('dplyr')
library(dplyr)

retainSummary <-
  dRetain %>%
  filter(SEX_CODE=='F') %>%
  group_by(FISCAL_YEAR) %>%
  summarise(students = n_distinct(STUDENT_ID),
            retained = n_distinct(STUDENT_ID[RETAINED == 'Y'])) %>%
  mutate(retainPercent = retained / students)  %>%
  arrange(desc(FISCAL_YEAR))


retainSummary

```

## Joining Data

How has retention varied by first generation status?

```{r}


dRetainFull <- left_join(dRetain,dFAFSA, by = 'STUDENT_ID')
dRetainFull <- left_join(dRetain,dFAFSA, by = c('STUDENT_ID','FISCAL_YEAR'))

length(dRetainFull$STUDENT_ID[is.na(dRetainFull$FIRST_GEN)])


```


# Visualizing Data

How has retention varied by first generation status (Continued)

```{r}

#I typically put the loading of packages all near the top of the document so I can use it throughout. Have it here to illustrate what packages I am using.
library(ggplot2)
library(scales)

dPlotOverall <-
  dRetainFull %>%
  group_by(FISCAL_YEAR) %>%
  summarise(students = n_distinct(STUDENT_ID), retained = n_distinct(STUDENT_ID[RETAINED == 'Y'])) %>%
  mutate(pRetain = retained / students)

pOverall <-
  ggplot(data = dPlotOverall,aes(x = FISCAL_YEAR, y = pRetain)) +
  geom_line(group = 1, color = rgb(244, 121, 32, maxColorValue = 255), linewidth = 1.2) +
  geom_point(size = 2) +
  geom_text(aes(label = paste(percent(pRetain,accuracy = .1),'\n(', retained,' of ',comma(students),')', sep = '')),vjust = -.3) +
  scale_y_continuous(limits = c(0,1), labels = percent_format()) +
  labs(title = 'Fall-to-Fall Retention by Year', x = '', y = 'Percent Retained') +
  theme_bw() +
  theme(axis.text = element_text(size = 12))
  

pOverall

```

```{r, fig.height=8}

dPlotOverall <-
  dRetainFull %>%
  filter(is.na(FIRST_GEN) == FALSE) %>%
  group_by(FISCAL_YEAR, FIRST_GEN) %>%
  summarise(students = n_distinct(STUDENT_ID), retained = n_distinct(STUDENT_ID[RETAINED == 'Y'])) %>%
  mutate(pRetain = retained / students) %>%
  mutate(FIRST_GEN_DESC = ifelse(FIRST_GEN == 'Y','First Gen','Not First Gen'))



pFirstGen <-
  ggplot(data = dPlotOverall,aes(x = FISCAL_YEAR, y = pRetain)) +
  geom_col(fill = rgb(244, 121, 32, maxColorValue = 255), linewidth = 1.2) +
  geom_text(aes(label = paste(percent(pRetain,accuracy = .1),'\n(', retained,' of ',comma(students),')', sep = '')),vjust = -.3) +
  scale_y_continuous(limits = c(0,1), labels = percent_format()) +
  labs(title = 'Fall-to-Fall Retention by Year', x = '', y = 'Percent Retained') +
  theme_bw() +
  theme(axis.text = element_text(size = 12)) +
  facet_wrap(~FIRST_GEN_DESC, nrow = 2, ncol = 1)

pFirstGen


```


# Statistics

Probably wont have time to cover during the session. Brief example of regression analysis below though.

```{r}

dModel <-
  dRetainFull %>%
  filter(is.na(FIRST_GEN) == FALSE & is.na(PELL_ELIGIBLE) == FALSE)

#Linear Regression
mRetentionLinear <- lm(data = dModel, RETAIN_CODE ~ HS_GPA + SEX_CODE + IPEDS_RACE_CODE + FIRST_GEN + PELL_ELIGIBLE)

summary(mRetentionLinear)

#Logistic Regression
mRetentionLogistic <- glm(data = dModel, RETAIN_CODE ~ HS_GPA + SEX_CODE + IPEDS_RACE_CODE + FIRST_GEN + PELL_ELIGIBLE, family = "binomial")
summary(mRetentionLogistic)


```




